Rapid-response Bayesian sky localization for electromagnetic follow-up of gravitational-wave candidates

### 2013-06-21 ms.tex
Margin note about mis-definition of horizon distance
Described hierarchical refinement

Assuming circular orbits and no spin precession, we can write the \ac{GW} signal in each detector as a linear combination of two basis waveforms $h_0$ and $h_{\pi/2}$. For nonprecessing systems, $h = h_0$ and $h_{\pi/2}$ are approximately ``in quadrature'' in the same sense as the sine and cosine functions, being nearly orthogonal and out of phase by ${\pi/2}$ at all frequencies. In other words, $h_{\pi/2}(\omega) = i h_0(\omega)$ for all $\omega$. This linear combination isolates all dependence on the extrinsic parameters $\boldsymbol\theta_\mathrm{ex}$ into the coefficients and all dependence on the intrinsic parameters $\boldsymbol\theta_\mathrm{in}$ into the basis waveform,
%
\begin{eqnarray}\label{eq:full-signal-model}
	\fl h_i(\omega; \boldsymbol\theta) &=& e^{-i\omega t_\oplus}\frac{D_{1,i}}{D_\mathrm{L}}
	\left(\begin{array}{cc}
		F_{+,i}(\alpha, \delta, t_\oplus) \\
		F_{\times,i}(\alpha, \delta, t_\oplus)
	\end{array}\right)^\intercal
	\left(\begin{array}{cc}
		\cos{2\psi} & -\sin{2\psi} \\
		\sin{2\psi} & \cos{2\psi}
	\end{array}\right)
	\left(\begin{array}{cc}
		\frac{1}{2}\left(1+u^2\right) & 0 \\
		0 & u
	\end{array}\right) \nonumber\\\fl &\cdot&
	\left(\begin{array}{cc}
		\cos{2\phi_c} & -\sin{2\phi_c} \\
		\sin{2\phi_c} & \cos{2\phi_c}
	\end{array}\right)
	\left(\begin{array}{cc}
		1 \\
		i
	\end{array}\right) h(\omega; \boldsymbol\theta_\mathrm{in}).
\end{eqnarray}
%
This linear combination matches the conventions used in \cite{PhysRevD.83.084002}, and is the frequency-domain equivalent of their Equation~(2.8). The quantities $F_{+,i}$ and $F_{\times,i}$ are the dimensionless detector antenna factors, and depend on the orientation of detector $i$ as well as the sky location and sidereal time of the event, and are derived in~\cite{PhysRevD.63.042003}. The quantity $D_{1,i}$ is a fiducial horizon distance for detector $i$,
%
\begin{equation}\label{eq:horizon}
D_{1,i} = 1 / \sigma_i, \qquad {\sigma_i}^2 = \int_0^\infty \frac{\left|h(\omega; \mathbf \theta)\right|^2}{S_i(\omega)} \,\mathrm{d}\omega.
\end{equation}

@@

The overall arrival time $t_\oplus$ can be marginalized out analytically, giving the \ac{TOA}-only marginal posterior
%
\begin{equation}
    p(\mathbf{n} | \hat{\tau}_1, \hat{\tau}_2, \dots, \hat{\tau}_N) \propto \exp\left[
        -\frac{1}{2} \sum_{i=1}^N \frac{\left( \tilde{\tau}_i - \langle \tilde{\tau} \rangle \right)^2}{\left(\hat{\rho_i} / \omega_{\mathrm{rms},i}\right)^2}
    \right]
\end{equation}
%
where
%
\begin{equation}
    \langle \tilde{\tau} \rangle =
    \left(\sum_{i=1}^N \frac{{\tilde{\tau}_i}^2}{\left(\hat{\rho_i} / \omega_{\mathrm{rms},i}\right)^2}\right)\left(\sum_{i=1}^N \frac{1}{\left(\hat{\rho_i} / \omega_{\mathrm{rms},i}\right)^2}\right)^{-1}.
\end{equation}

@@

\begin{figure}
    \caption{\label{fig:radial_integrand}Illustration of initial subdivisions for distance integration scheme. Distance increases from left to right. In the print version, the left\nobreakdashes-hand tail, the left\nobreakdashes- and right\nobreakdashes-hand sides of the maximum likelihood peak, and the right\nobreakdashes-hand tail, are colored cyan, red, green, and blue, respectively.}
    \begin{center}
        \includegraphics{radial_integrand}
    \end{center}
\end{figure}

@@

\section{Hierarchical refinement}

\marginpar{Talk about adaptive selection of HEALPix resolution and of pixels on which to evaluate \ac{SNR} marginal posterior. Something to think about: this is logically easier to explain last, but chronologically makes more sense first.}

@@

Some of the results in this paper have been derived using HEALPix \cite{healpix}.

>>>>

Assuming circular orbits and no spin precession, we can write the \ac{GW} signal in each detector as a linear combination of two basis waveforms $h_0$ and $h_{\pi/2}$. For nonprecessing systems, $h = h_0$ and $h_{\pi/2}$ are approximately ``in quadrature'' in the same sense as the sine and cosine functions, being nearly orthogonal and out of phase by ${\pi/2}$ at all frequencies. In other words, $h_{\pi/2}(\omega) = i h_0(\omega)$ for all $\omega$. This linear combination isolates all dependence on the extrinsic parameters $\boldsymbol\theta_\mathrm{ex}$ into the coefficients and all dependence on the intrinsic parameters $\boldsymbol\theta_\mathrm{in}$ into the basis waveform,
%
\begin{eqnarray}\label{eq:full-signal-model}
	\fl h_i(\omega; \boldsymbol\theta) &=& e^{-i\omega t_\oplus}\frac{D_{1,i}}{D_\mathrm{L}}
	\left(\begin{array}{cc}
		F_{+,i}(\alpha, \delta, t_\oplus) \\
		F_{\times,i}(\alpha, \delta, t_\oplus)
	\end{array}\right)^\intercal
	\left(\begin{array}{cc}
		\cos{2\psi} & -\sin{2\psi} \\
		\sin{2\psi} & \cos{2\psi}
	\end{array}\right)
	\left(\begin{array}{cc}
		\frac{1}{2}\left(1+u^2\right) & 0 \\
		0 & u
	\end{array}\right) \nonumber\\\fl &\cdot&
	\left(\begin{array}{cc}
		\cos{2\phi_c} & -\sin{2\phi_c} \\
		\sin{2\phi_c} & \cos{2\phi_c}
	\end{array}\right)
	\left(\begin{array}{cc}
		1 \\
		i
	\end{array}\right) h(\omega; \boldsymbol\theta_\mathrm{in}).
\end{eqnarray}
%
This linear combination matches the conventions used in \cite{PhysRevD.83.084002}, and is the frequency-domain equivalent of their Equation~(2.8). The quantities $F_{+,i}$ and $F_{\times,i}$ are the dimensionless detector antenna factors, and depend on the orientation of detector $i$ as well as the sky location and sidereal time of the event, and are derived in~\cite{PhysRevD.63.042003}. The quantity $D_{1,i}$ is a fiducial horizon distance for detector $i$,
%
\marginpar{This isn't what we usually call the horizon distance; the horizon distance is where face-on inspirals produce SNR=8.}%
%
\begin{equation}\label{eq:horizon}
D_{1,i} = 1 / \sigma_i, \qquad {\sigma_i}^2 = \int_0^\infty \frac{\left|h(\omega; \mathbf \theta)\right|^2}{S_i(\omega)} \,\mathrm{d}\omega.
\end{equation}

@@

The overall arrival time $t_\oplus$ can be marginalized out analytically, giving the \ac{TOA}\nobreakdashes-only marginal posterior
%
\begin{equation}\label{eq:toa-posterior}
    p(\mathbf{n} | \hat{\tau}_1, \hat{\tau}_2, \dots, \hat{\tau}_N) \propto \exp\left[
        -\frac{1}{2} \sum_{i=1}^N \frac{\left( \tilde{\tau}_i - \langle \tilde{\tau} \rangle \right)^2}{\left(\hat{\rho_i} / \omega_{\mathrm{rms},i}\right)^2}
    \right]
\end{equation}
%
where
%
\begin{equation}
    \langle \tilde{\tau} \rangle =
    \left(\sum_{i=1}^N \frac{{\tilde{\tau}_i}^2}{\left(\hat{\rho_i} / \omega_{\mathrm{rms},i}\right)^2}\right)\left(\sum_{i=1}^N \frac{1}{\left(\hat{\rho_i} / \omega_{\mathrm{rms},i}\right)^2}\right)^{-1}.
\end{equation}

\subsection{Adaptive resolution}

The \ac{TOA}\nobreakdashes-only posterior in \Eref{eq:toa-posterior} is numerically trivial to evaluate because it involves no numerical integrals. However, as we will see in the next section, the \ac{SNR}\nobreakdashes-only posterior is numerically expensive.

We evaluate the \ac{TOA}\nobreakdashes-only posterior at the centers of \ac{HEALPix} pixels at the lowest possible resolution. Then, we increase the resolution until 99.99\% of the \ac{TOA}\nobreakdashes-only posterior is contained within at least 3072 pixels. The numerically expensive \ac{SNR} integrals are done only for the pixels that comprise the 99.99th percentile at this resolution. All other pixels are set to zero.

@@

\begin{figure}
    \caption{\label{fig:radial_integrand}Illustration of initial subdivisions for distance integration scheme. Distance increases from left to right. In the color version, the left\nobreakdashes-hand tail, the left\nobreakdashes- and right\nobreakdashes-hand sides of the maximum likelihood peak, and the right\nobreakdashes-hand tail, are colored cyan, red, green, and blue, respectively.}
    \begin{center}
        \includegraphics{radial_integrand}
    \end{center}
\end{figure}

@@

@@

Some of the results in this paper have been derived using \ac{HEALPix} \cite{healpix}.
